---
title: "Survival Analysis Project - José Lise"
output:
  pdf_document: default
  html_notebook: default
---

# Employee turnover study using survival analysis 

The goal of this project is to carry out the analysis of employee churn. 
We will work with a dataset available on [github](https://github.com/ludovicbenistant/Management-Analytics/blob/master/HR/HR.csv). The dataset description is also available on [kaggle](https://www.kaggle.com/c/employee-churn-prediction/data), althouth it cannot be freely downloaded from there.  

Here is the dataset description: it's a csv file with 15 000 rows (1 row for column headers and 14 499 rows for employees information) and 10 columns.  
  
Here are the columns description:    
+ satisfaction_level - Employee satisfaction level    
+ last_evaluation - Last evaluation score    
+ number_project - Number of projects assigned to  
+ average_monthly_hour - Average monthly hours worked  
+ time_spend_company - Time spent at the company in years. Between 2 and 10.  
+  Work_accident - (1= Yes, 0 = No) - Whether they have had a work accident  
+ left - Whether or not employee left company (1 = Yes, 0 = No). This is the censoring information.   
+ promotion_last_5year - (1 = Yes, 0 = No) - Whether they have had a promotion in the last 5 years  
+ sales - Department name (not just sales)  
+ salary - Salary category  

## Data processing 
### Loading data
The following cell load the data from the csv file and assign column types numeric or character according the data stored. 


```{r include=FALSE}
library(tidyverse)
raw <- read_delim("HR.csv",  delim=";", col_names=TRUE,
           col_types = cols(satisfaction_level = 'd',
                            last_evaluation = 'd',
                            number_project = 'd',
                            average_montly_hours = 'd',
                            time_spend_company = 'd',
                            Work_accident = 'd',
                            left='d',
                            promotion_last_5years = 'd',
                            sales = 'c',
                            salary = 'c'))
```

### Data transformation
In the following cell, we transform the data to assign factors to obvious categorical variables (promotion_last_5years, Work_accident, sales and salary).   
In addition, we also manage number_project as a categorical variable even it can be seen as an integer numeric variable. This decision was taken after a first exploratory analysis of the data. 

```{r}
# Assign factor level for promotion_last_5years, Work_accident, number_project, sales and salary
# Rename sales column to department (it's less confusing) 
h <- mutate(raw,
            Work_accident = factor(Work_accident, levels=c('0','1'), labels = c('no', 'yes')),
            promotion_last_5years = factor(promotion_last_5years,levels=c('0','1'),
                                           labels = c('no', 'yes')),
            number_project = factor(number_project),
            sales = factor(sales),        
            salary = factor(salary))

h <- h %>% rename(department = sales)
```

### Exploratory analysis

```{r}
summary(h)
```

We see that there are no missing values. The time variable that we will use for our survival analysis is **time_spend_company**.It's a time in years and goes from 2 to 10 years. The censoring variable is called *left*: it's 1 if the employee has left the company and 0 otherwise.   
**satisfaction_level** and **last_evaluation** are decimal numbers between 0 and 1.  
**number_project** is an integer number in the range 2 to 7 managed as a categorical variable.   
**average_montly_hours** is an integer number in the range 96 to 310. 
There are 2 binaries categorical variables: **Work_accident** and **promotion_last_5years**.
And the last 2 variables *department* and *salary* are also categorical but with at least 3 levels. 


```{r}
hist(h$time_spend_company)

```

We see that employees start leaving the company after 2 years. The peak duration is 3 years. 
The number of employees leaving the company decrease from 3 years to 6 years. And from 6 years to 10 years there is almost no change. 



Check how many employees left the company.

```{r}
table(h$left)

```
Out of the 14 499 employees, 3571 left. 

### Satisfaction level, Last evaluation and Average monthly hours

Global histograms
```{r}
par(mfrow=c(1,3))
hist(h$satisfaction_level, col = 'blue', main="Global satisfaction level")
hist(h$last_evaluation, col = 'blue', main="Global last evaluation")
hist(h$average_montly_hours, col = 'blue', main="Global average montly hours")
```

For the employees who left.
```{r}
par(mfrow=c(1,3))
hist(h$satisfaction_level[h$left == 1], col='red', main='Left satisfaction level')
hist(h$last_evaluation[h$left == 1], col='red', main='Left last evaluation')
hist(h$average_montly_hours[h$left == 1], col='red', main='Left average montly hours')
```
The satisfaction histogram for employee who left shows that they are very unsastified or neutral. And a significant part was above average satisfied.   
For the last evalation, most of the employees who left were average or below average. And a signicant amount was above average.  
For the working hour, most of the employees who left were in the lower bucket( 150 hours or less). 



### Work accident and promotion 
Global barplots

```{r}
par(mfrow=c(1,2))
barplot(prop.table(table(h$Work_accident)) , main="Global work accident")
barplot(prop.table(table(h$promotion_last_5years)), main="Global promotion last 5years")
```

Left barplots

```{r}
par(mfrow=c(1,2))
barplot(prop.table(table(h$Work_accident[h$left == 1])), main="Left work accident")
barplot(prop.table(table(h$promotion_last_5years[h$left == 1])), main="Left promotion last 5years")
```
The proportion of work accident is lower in the population that left. And the proportion of promotion in the last 5 years is also lower for the employees that left.  


### Department, salary and number of projects
Global barplots

```{r}
par(mfrow=c(1,3))
barplot(table(h$department) , main="Global department")
barplot(table(h$salary), main="Global salary")
barplot(table(h$number_project), main="Global Number of projects")
```
Left barplots


```{r}
par(mfrow=c(1,3))
barplot(table(h$department[h$left == 1]), main="Left department")
barplot(table(h$salary[h$left == 1]), main="Left salary")
barplot(table(h$number_project[h$left == 1]), main="Left Number of projects")
```
From the plot above, we can already think that there is a higher risk (of leaving company)  for employees with low salaries.


## Kaplan-Meier estimator
We draw below the Kaplan-Meier survival curve. 

```{r}
library(survival)
St <- survfit(Surv(time_spend_company,left) ~ 1, data = h)
St
plot(St, main="Kaplan-Meier estimator", xlab="Years", ylab="Survival probability")
```

The median value returned by the survfit function is 6 years. Therefore the average lifetime at the company for an employee is 6 years. The confidence interval for the median is (5,6) in years. We also notice that the survival probability is constant after 6 years. 



Survival curve by salary 
```{r}
Sts <- survfit(Surv(time_spend_company,left) ~ salary, data = h)
Sts
plot(Sts, col=c('black','red','blue'),main="Survival curve by salary", xlab="Years", ylab="Survival")
legend("bottomleft", legend=levels(h$salary),
       col=c("black", "red", "blue"), lty=1, cex=0.8)

```
We see that the median for salary=high is undefined, because for this category the survival function is always above 0.5. The group with higher risk is salary=low group. The median for this group is 5 years, whereas it's 6 years for the medium group. 


### Comparing groups
We will compare the groups for categorical varaiables using the log rank test: Work_accident, promotion_last_5years, number_project, department, and salary.

```{r}
test_variable <- function(var_name) {
  h$x <- h[[ var_name ]]
  survdiff(Surv(time_spend_company,left) ~ x, data = h)
}

logrank_tests <-
  tibble(variable = c("Work_accident", "promotion_last_5years","number_project", "department", "salary")) %>%
  mutate(obj = map(variable, test_variable),
         tab = map(obj, broom::glance)) %>%
  unnest(tab)
logrank_tests


#survdiff(Surv(time_spend_company,left) ~ salary, data = h)

```

The table above lists the p-values for the log-rank test for the 5 selected features. And they all are zero.Therefore we reject the null hypothesis for all the variables and there is a significant difference between the groups inside each variable.  



## Data Modeling and Machine Learning

### Split the data randomly into a training and a testing set


Reshape the data for the modeling section 

```{r}
hm <- h 
hm <- select(hm, -c(time_spend_company, left))
summary(hm) 
```

```{r}
y <- Surv(h$time_spend_company, h$left)

```

### Creating Training and Testing sets

We will split the 14499 rows in 2 sets: 10000 row for training and 4999 for testing.

```{r}
set.seed(1234) # >To make the results replicable
i.training <- sample.int(nrow(hm), size = 10000, replace = FALSE)
i.testing <- setdiff(seq_len(nrow(hm)), i.training)

h.training <- hm[i.training,, drop = FALSE]
y.training <- y[i.training,, drop = FALSE]

h.testing <- hm[i.testing,, drop = FALSE]
y.testing <- y[i.testing,, drop = FALSE]
```

### Cox regression

```{r}
fit <- coxph(y.training ~ ., data = h.training)
summary(fit)
```
Cox Hazard Ratio Analysis:
Most of the coefficients are significants (P-values very small or less than 5%) except for some of the department levels. 

The variable with the more significant coefficient is number_project3. And this coefficient is negative and therefore reduce the risk: Working on 3 projects reduce the risk by a factor 49 CI(36, 66) compared to working on 2 projects (which is the reference level). 
And we see also that working on 4 projects decrease the risk by a factor 11.9 CI( 10.3, 13.7)compared to working on 2 projects.And the risk is decreased  by a factor 7 CI(6.2,8), 4.9 CI(4.2, 5.6) and 3.2 CI(2.6, 3.9) for respectively 5, 6 and 7 projects. 

The next most significant feature in the satisfaction level: An increase by one unit (which is the full range) reduces the risk by a factor 3.95 CI(3.3, 4.8) .

Then in the significance level we have the salary impact: risk is increased by a factor 3.38 CI(2.6, 4.4) from Salary_high to Salary_low, and by a factor 2.4 CI(1.8, 3.2) from salary_high to salary_medium. 

Then we have in terms of significance level: promotion_last_5yearsyes. A promotion in the last 5 years reduce the risk by a factor 3.5 CI (2,6.3) compared to no promotion.

Next significant variable is Work_accidentyes: The risk is reduced by a factor 2.9 CI( 2.4,3.5) compared to Work_accidentNo. This is counter intuitive, but looking at the exploratory plot, the proportion of Work_accidentyes is lower for people who left. 

There is not much difference in hazard ratio coefficient for the department variable. The higher evidence, is that risk for R&D departent compared to accounting is lower by a factor 1.6 CI(1.2, 2.2). 



### Model 1 - AIC Step
In this section we will the step function to select the model with the lower AIC. 

```{r}
fit.aic <- step(fit)
summary(fit.aic)
b.aic <- coef(fit.aic)

```
The AIC step function finds that the best model (lower AIC)  is the model with all the variables. 

The Concordance is 0.911 and therefore is quite high.

#### Models testing: AIC
In the following section, we will test the model selected by AIC. 
We apply the model previously fitted to the testing dataset, to get the scoreAIC.testing. 

```{r}
scoreAIC.testing <- predict(fit.aic, newdata = h.testing)
hist(scoreAIC.testing)

```
The score is in the range (-4, 3). 
y.testing is a censored variable and scoreAIC.testing is assumed to be a continuous predictor of this variable.
A continuous predictor vs a right-censored time-to-failure outcome: therefore we can use Cox regression.

```{r}
summary(coxph(y.testing ~ scoreAIC.testing))

```
Coefficient is positive(1.04) therefore higher score means higher risk. 
The hazard ratio is 2.83 with confidence interval  (2.67, 2.98).
And as the p-value is small, this means this coefficient is significant. 

We also notice that the concordance is very high: 0.913 with a standard error of 0.005. 
This model score is very good, but we use all the features provided. 
Let's see if we can reduce the number of features while keeping the same kind of performance. 


### model 2. elastic net

```{r include=FALSE}
library(glmnet)
# we remove the intercept to get a matrix with only the covariates
X <- model.matrix( ~ . -1, data = h.training)
fit.cv10 <- cv.glmnet(X, y.training, family = "cox")
```


plotting

```{r}
plot(fit.cv10)
```
We select Lambda.1se the less conservative lambda, which is also the lambda with the lower number of coefficients. 


```{r}
b <- coef(fit.cv10, s = "lambda.1se")
sum(b != 0)
b.enet <- b[b != 0]
```
Lambda.1se has 15 non null coefficients.


```{r}
round(b[abs(b) > 1e-3 ], digits = 3)
```

#### Model testing: elastic net

```{r}
X.testing <- model.matrix( ~ . -1 , data = h.testing)
score.testing <- predict(fit.cv10, newx = X.testing, s = "lambda.1se")
hist(score.testing)
```
Higher score means higher risk and then lower survival.
The score is in the range (-3, 3). 

y.testing is a censored variable and score.testing is assumed to be a continuous predictor of this variable.
A continuous predictor vs a right-censored time-to-failure outcome: we can use Cox regression.

```{r}
summary(coxph(y.testing ~ score.testing))
```
Coefficient is positive (1.22) therefore higher score means higher risk. 
The hazard ratio is 3.38 with confidence interval  (3.189, 3.594).
And as the p-value is small, this means this coefficient is significant. 

And the concordance is the same as with the previous model: 0.913 with a slightly higher SE of 0.006. 

And this model is smaller than the other one we will carry out some additional analysis with it. 

#### Low and high categories risk 
We can split the scores into 2 categories, and compare employees with 'low' vs 'high' score:

```{r}
x_risk <- ifelse(score.testing <= median(score.testing), "low", "high")
table(x_risk)
```


```{r}
fit.KM <- survfit(y.testing ~ x_risk, conf.type = "log-log")
plot(fit.KM, col=c('red','blue'),main="Survival curve by risk level", xlab="Years", ylab="Survival")
legend("bottomleft", legend=c("high", "low"),
       col=c( "red", "blue"), lty=1, cex=0.8)
```

Using the log rank test to check the difference between low and high risk. 
```{r}
survdiff(y.testing~ x_risk)

```
The p-value is zero, therefore the difference between high risk and low risk is significant. 


```{r}
fit.KM
```

We see that the median is 5 years CI(5,5) for low risk. And the median is undefined for high risk (the survival is always above 0.5). 

Survival at 4 years:
```{r}

summary(fit.KM, time = 4)

```
At 4 years, the survival is 0.56 CI(0.53, 0.58)  for the high risk population and 0.97 CI(0.96, 0.9) for the low risk. 


AUC index at 4 years

```{r}
library(survivalROC)
ROC <- survivalROC(Stime = y.testing[, 1],
                   status = y.testing[, 2],
                   marker = score.testing,
                   cut.values = quantile(score.testing, prob = 0:100/100),
                   predict.time = 4,
                   method = "KM")
ROC$AUC


```
The AUC at 4 years is 0.82, therefore quite high. 




## Conclusion

We analyzed the employees turnover of a company using survival analysis modeling over a period of 10 years. 
After an exploratory analysis, we computed the Kaplan-Meier estimator for survival probability. Then we identified that the median survival at the company is 6 years: the expected time spent at the company is 6 years. 
Then we use the log rank test to check if the levels in the 5 categorical features have an impact on the survival. The log rank test was positive for all the categorical variables. 

We then splitted the initial 14499 rows dataset in one training dataset of 10000 rows and a testing dataset of 4999 rows, before to carry out the modeling part. 

We first did ran a cox regression on the training set and idenfied hazard ratio coefficients with the most impact on the risk: the number of projects. And specifically between 2 and 3 projects there is a factor of 44. 

We then tried to do model selection using AIC and GLMNET. 
AIC returned the full model as the best model and GLMNET returned a model with 15 coefficients out of 20. 

We then carried out some predictions using the test dataset and use cox regression again to assess the accuracy of the models. Both model gave a very high concordance of 0.913. 



